The Power of Unbiased Recursive Partitioning:\\ A Unifying View of CTree, MOB, and GUIDE

### 2019-05-12 pkg/partykit/inst/guideliketest/draft/draft.Rnw
rewrote sections 5.1. and 5.2. and edited captions of section 5.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2553 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 7d65c57349137404aee06a59161a71fced4ca478

Even though CTree, MOB, and GUIDE differ in the specific tests they apply, their approaches for split
variable selection follow the same basic structure as explained in Section~\ref{sec:generic_alg}. 
In fact, the tests can be embedded in a unifying conceptual framework that yields the
different tests by combining various building blocks.
These mostly differ in the way the model for the dependent variable on the one hand and
the splitting variables on the other are prepared or transformed:
%
\begin{itemize}
  \item \emph{Goodness-of-fit measure:}\\
        Different variations of the disaggregated, observation-wise goodness-of-fit measure of the 
        model for the response $Y$ and possible covariates $X$ can be considered. Either residuals
	$r(Y,X,\hat{\beta})$ can be used as proposed for GUIDE or model scores $s(Y,X,\hat{\beta})$
        as proposed for MOB. All three algorithms can, in principle, use both goodness-of-fit
	measures though, which is probably brought out most clearly in the CTree algorithm that
        explicitly allowed for different transformations $h(Y)$ in its original description already.
  \item \emph{Binarization of residuals/scores:}\\
        Rather than testing independence between the split variables and the residuals/scores
	tmeselves, it is possible	
	to binarize residuals/scores at $0$ so that only their signs are assessed (as proposed for GUIDE).
  \item \emph{Categorization of split variables:}\\
        Similarly, the split variables can also be categorized (for testing only). This was proposed for
	GUIDE, employing binning at the quartiles yielding four categories of approximately
	equal size.
\end{itemize}
%
The three algorithms combine these building blocks in different ways as shown in Table~\ref{tab:buildingblocks}:
When applying CTree for unbiased model-based recursive partitioning it has been suggested to use
the model scores without binarization and assess their association with the untransformed split
variables using a conditional inference test. This is similar to a squared correlation test statistic.
MOB also employs the scores without
binarization and maximally selects a score statistic over all potential split points in the
split variable. GUIDE employs the binarized residuals and assesses their association with
the categorized split variable in a classical (unconditional) $\chi^2$ test.

@@

\subsection{Evaluation}
\label{sec:evaluation}
The testing strategies are evaluated over a stepwise increasing effect size $\delta$, 
on 100 replications per step each consisting of 250 observations. 
To compare the performance of the evaluated testing strategies for each step 
in the first scenario (stump with 1 split) the following criteria are considered:
the $p$-values pertaining only to the true split variable $Z_1$;
and the proportion of replications for which the $p$-value of $Z_1$
is the lowest and significant at 5\% level (i.e., where $Z_1$ would be
selected for splitting in a pre-pruning approach).
For the second scenario (tree with 2 splits) the adjusted Rand index~(ARI) is calculated 
as a measure of similarity between the true tree structure and the fitted model tree.

@@

@
\caption{\label{fig:scores_stump}Proportion of replications where the true split variable $Z_1$ is
selected over increasing effect size $\delta$ (100 replications of 250 observations per step),
once for a true split point $\xi = 0$ (top row) and once for $\xi = 0.8$ (bottom row). 
Each column represents one scenario for the coefficients: 
varying intercept $\beta_0$ (left), varying slope parameter $\beta_1$ (middle) and 
both coefficients varying (right). The performances of four testing strategies are compared: 
CTree, MOB, GUIDE, GUIDE+scores.}
\end{figure}

Figure~\ref{fig:scores_stump} shows the evaluation results for the three different
scenarios (varying only the intercept $\beta_0$, only the slope parameter $\beta_1$
or both of them) once for a true split point $\xi=0$ and once for $\xi=0.8$. 
As explained before, the tests are applied as significance tests with significance 
level $\alpha=0.05$. Therefore, the proportion of replications in which the true split variable $Z_1$
is detected with a $p$-value smaller than $\alpha$ is compared.
(To show that in general the drawn conclusions do not change if no significance level is included 
the same setting is evaluated again without any significance level and the results are presented
in Appendix~\ref{app:significance}.)

For a central split point ($\xi=0$, upper panels) all strategies perform similarly if the 
intercept varies (left and right panels) where CTree and MOB are only slightly ahead. However, 
this advantage increases if the slope parameter varies and the intercept is kept at a fixed value 
(middle panel). In particular, the original strategy of GUIDE fails to detect any changes 
in $\beta_1$. 
This is due to the fact that GUIDE only includes residuals, i.e., only the scores 
regarding the intercept, such that changes in the slope parameter are not detected unless the 
corresponding scores are also included as in GUIDE+scores. This disadvantage of using 
only scores has already been elaborated in the literature of structural change
analysis \citep[see e.g.][]{Ploberger+Kramer:1992} and is also discussed 
in Section~\ref{sec:discussion}.

Looking at the results for the split point being close to the upper edge of the considered
interval ($\xi=0.8$, lower panels) all of the applied strategies perform worse than for the
central split point. In this case MOB is clearly ahead leading to the best results. CTree 
performs second best while both GUIDE versions struggle to detect the correct split variable 
even for a high effect size. 
(In order to show that the here observed advantage of MOB over CTree is mainly due to the 
abrupt shift in the model parameters $\beta_0$ and $\beta_1$ an additional evaluation is 
presented in Appendix~\ref{app:contbeta} where these parameters change continuously.)

Altogether these results clearly illustrate that overall the decision to include all available
scores when testing for dependencies in order to select split variables is advantageous over
considering only residuals, i.e., the scores regarding only the intercept, as changes not only in
the mean but also in all other parameters can be detected in that way. This is particularly obvious
when comparing GUIDE and GUIDE+scores. Moreover, from the fact that these two version are both
performing worse than MOB and CTree it can be deduced that categorizing split variables and
binarizing scores has a negative effect on the ability to detect the right split variable. 
To elaborate these effects more in detail the results for all possible combinations of the 
building blocks are discussed in a full factorial 3-way analysis.

\subsection{Full factorial analysis}\label{sec:3way}
In Figure~\ref{fig:3way} the effects of the building blocks are illustrated for the most general
case where the intercept and slope parameter are both varying. In the left panel the true split point is in the center, i.e., $\xi=0$ and an effect size of $\delta = 0.3$ is considered. 
This choice has been made based on the results shown in Figure~\ref{fig:scores_stump} 
in order to make the differences particularly well observable.
In the right panel the true split point is close to the upper end of the considered interval
($\xi=0.8$) requiring a higher effect size $\delta = 1$ in order to visualize results properly.
(Additional evaluations for increasing $\delta$ as in Section~\ref{sec:scores} 
can be found in Appendix~\ref{app:increasingdelta}).
The power of the different combinations is measured by the mean $p$-value for the true split 
variable~$Z_1$.
The possible levels for each of the three building blocks are listed in Table~\ref{tab:levels}.
In total 12 combinations are applied including the original testing strategies CTree, MOB, and GUIDE
and variations of them. A more detailed overview of all 12 combinations can be found in 
Appendix~\ref{app:combinations}, Table~\ref{tab:combinations_appendix}.

@@

@@

#par(mar = c(3,5,1,5.7))
#with(d00, interaction.plot(res_scores, interaction(cat, bin), pval_z1, ylim = c(0,1),
#                           ylab = expression("Mean"~p~"-value for"~Z[1])))
#with(d00, interaction.plot(interaction(cat, bin), res_scores, pval_z1))
#with(d00, interaction.plot(cat, interaction(bin, res_scores), pval_z1))
#with(d08, interaction.plot(res_scores, interaction(cat, bin), pval_z1, ylim = c(0,1)))
#with(d08, interaction.plot(interaction(cat, bin), res_scores, pval_z1))
#with(d08, interaction.plot(cat, interaction(bin, res_scores), pval_z1))
@
\caption{\label{fig:3way} Mean $p$-values corresponding to the true split variable $Z_1$
over different levels of the building block \textit{Categorization}. 
Each line represents
a combination of the levels of the other two building blocks \textit{Binarization} and 
\textit{Residuals vs.\ Scores}.
Left panel: true split point $\xi=0$ and effect size $\delta = 0.3$.
Right panel: true split point $\xi=0.8$ and effect size $\delta = 1$.}
\end{figure}

The results presented in Figure~\ref{fig:3way} clearly show that binarizing scores/residuals
reduces the power of the testing strategy as it leads to higher $p$-values corresponding
to the true split variable $Z_1$. 
This effect can be observed for a central split point (left panel, $\xi=0$) as well as for a
a split point closer to upper end of the considered interval (right panel, $\xi=0.8$).
However, the effect of the building block \textit{Categorization} varies over these two settings.
For a central split point all three levels perform similarly. The linear selection of split variables
is slightly ahead even though the true parameter function is a step function. 
On the contrary, clear differences can be seen for the outward split point. Here, the maximal selection
performs best while the linear selection shows its disadvantage in detecting abrupt changes
and the categorized split variables lead to even higher $p$-values and thus lower power.
Overall, it depends on the situation whether a linear or a maximal selection of a split variable
is advantageous. However, categorizing possible split variables does not lead to any improvements
but even declines the performance of a testing strategy unless the true split point is close
to a break point used for the binarization.

@@

@
\caption{\label{fig:tree_ari}Adjusted Rand Index over increasing effect size $\delta$ 
(100 replications of 250 observations per step).
Each panel represents a different value of the true split point $\xi$ ($0$, $0.2$, $0.5$ and 
$0.8$ from left to right). The performances of four testing strategies are compared: CTree,
MOB, GUIDE, GUIDE+scores.}
\end{figure}

@@

\begin{figure}%[t]
\setkeys{Gin}{width=\linewidth}
<<prune_ari, fig=TRUE, echo=FALSE, height=4, width=8>>=
xyplot(ari ~ delta | xi, groups = ~ test, 
       data = prex, 
       type = "l", lty = c(2,2,2,2,1,1,1,1),
       lwd = 2,
       key =  list(text = list(levels(prex$test)[1:4]),
                   lines = list(lty = c(1,1,1,1), type = "l", pch = 1, cex = 0.8,
                                col = pal[c("CTree", "MOB", "GUIDE+scores", "GUIDE")], lwd = 2)),
       layout= c(4,1),
       scales=list(x=list(at=seq(1,11,2))),
       xlab = expression(delta),
       ylab = "Adjusted Rand Index",
       strip = strip.custom(factor.levels = c(expression(xi==0), 
                                              expression(xi==0.2), 
                                              expression(xi==0.5), 
                                              expression(xi==0.8)),
                                   bg = "gray90"),
       par.settings = par.settings
)
@
\caption{\label{fig:pruning_ari}Adjusted Rand Index over increasing effect size $\delta$ 
(100 replications of 250 observations per step).
Each panel represents a different value of the true split point $\xi$ ($0$, $0.2$, $0.5$ and 
$0.8$ from left to right). 
The performances of four testing strategies are compared: CTree, MOB, GUIDE, GUIDE+scores. 
The dashed lines represent the results from applying pre-pruning and the solid lines 
the results from applying post-pruning.}
\end{figure}

@@

For the evaluation of all 12 possible combinations of the presented building blocks
the original testing strategies CTree, MOB, and GUIDE have been applied together
with modified versions of them. The employed versions and the corresponding combination
of building blocks is listed in Table~\ref{tab:combinations_appendix}.
Each suffix indicates a change of the original form of the considered testing strategy.
In particular, it relates to the transformations of the data before applying a statistical
test as represented by the different building blocks. In that way,
``+res'' refers to using only residuals (instead of all scores),
``+scores'' to using all scores (instead of only residuals),
``+cat'' to applying a categorization of split variables,
``+bin'' to applying a binarization of scores/residuals.

>>>>

Even though CTree, MOB, and GUIDE differ in the specific tests they apply, their approaches for split
variable selection follow the same basic structure as explained in Section~\ref{sec:generic_alg}. 
In fact, the tests can be embedded in a unifying conceptual framework that yields the
different tests by combining various building blocks.
These mostly differ in the way the model for the dependent variable on the one hand and
the splitting variables on the other are prepared or transformed:
%
\begin{itemize}
  \item \emph{Goodness-of-fit measure:}\\
        Different variations of the disaggregated, observation-wise goodness-of-fit measure of the 
        model for the response $Y$ and possible covariates $X$ can be considered. Either residuals
	$r(Y,X,\hat{\beta})$ can be used as proposed for GUIDE or model scores $s(Y,X,\hat{\beta})$
        as proposed for MOB. All three algorithms can, in principle, use both goodness-of-fit
	measures though, which is probably brought out most clearly in the CTree algorithm that
        explicitly allowed for different transformations $h(Y)$ in its original description already.
  \item \emph{Binarization of residuals/scores:}\\
        Rather than testing independence between the split variables and the residuals/scores
	themselves, it is possible	
	to binarize residuals/scores at $0$ so that only their signs are assessed (as proposed for GUIDE).
  \item \emph{Categorization of split variables:}\\
        Similarly, the split variables can also be categorized (for testing only). This was proposed for
	GUIDE, employing binning at the quartiles yielding four categories of approximately
	equal size.
\end{itemize}
%
The three algorithms combine these building blocks in different ways as shown in Table~\ref{tab:buildingblocks}:
When applying CTree for unbiased model-based recursive partitioning it has been suggested to use
the model scores without binarization and assess their association with the untransformed split
variables using a conditional inference test. This is similar to a squared correlation test statistic.
MOB also employs the scores without
binarization and maximally selects a score statistic over all potential split points in the
split variable. GUIDE employs the binarized residuals and assesses their association with
the categorized split variable in a classical (unconditional) $\chi^2$ test.

@@

\subsection{Evaluation}
\label{sec:evaluation}
The testing strategies are evaluated over a stepwise increasing effect size $\delta$, 
on 100 replications per step each consisting of 250 observations. 
To compare the performance of the evaluated testing strategies for each step 
in the first scenario (stump with one split) the following criteria are considered:
the $p$-values pertaining only to the true split variable $Z_1$;
and the proportion of replications for which the $p$-value of $Z_1$
is the lowest and significant at 5\% level (i.e., where $Z_1$ would be
selected for splitting in a pre-pruning approach).
For the second scenario (tree with two splits) the adjusted Rand index~(ARI) is calculated 
as a measure of similarity between the true tree structure and the fitted model tree.

@@

@
\caption{\label{fig:scores_stump}
CTree, MOB, GUIDE, and GUIDE+scores evaluated on the first simulation scenario (stump with one split). 
Their performances are compared based on the proportion of replications where the true split 
variable $Z_1$ is selected over increasing effect size $\delta$ (100 replications of 
250 observations per step), once for a true split point $\xi = 0$ (top row) and once 
for $\xi = 0.8$ (bottom row). 
Left: varying intercept $\beta_0$. Middle: varying slope parameter $\beta_1$.
Right: both coefficients varying.}
\end{figure}

Figure~\ref{fig:scores_stump} shows that overall CTree and MOB (both using full 
model scores) perform clearly better than the original GUIDE version (using residuals 
only) as they accomplish to select the true split variable $Z_1$ in a higher proportion
of replications. Even though the advantage of CTree and MOB evens out in certain situations
such as for a high effect size (see, e.g., for $\xi=0$ and varying $\beta_0$), GUIDE is 
never ahead of them in any of the considered settings. Moreover, the positive effect of 
including all available scores can be observed particularly well in the top middle panel 
where GUIDE+scores detects a shift in the slope parameter $\beta_1$ as opposed to 
GUIDE. The fact that residual-based methods fail to detect changes that do not effect the
expected value has already been elaborated in the literature of structural change
analysis \citep[see, e.g.,][]{Ploberger+Kramer:1992} and is also discussed in 
Section~\ref{sec:discussion}. Moreover, the advantage of including full models scores 
clearly outweighs the cost involved which is displayed by the small gain of GUIDE compared 
to GUIDE+scores in case of a fixed slope parameter (left panels).

For a central split point ($\xi=0$, upper panels) all strategies perform similarly
well as long as the intercept varies (left and right panel) with CTree and MOB 
being only slightly ahead. However, for the split point being close to the upper 
edge of the considered interval ($\xi=0.8$, lower panels) the performance of all of 
the applied strategies decreases. 
While both GUIDE versions struggle to detect the correct split variable even for a 
high effect size, MOB is clearly ahead with CTree leading to the second best results.
(In order to show that the here observed advantage of MOB over CTree is mainly due to the 
abrupt shift in the model parameters $\beta_0$ and $\beta_1$ an additional evaluation is 
presented in Appendix~\ref{app:contbeta} where these parameters change continuously.)
From the fact that GUIDE and GUIDE+scores are both performing worse than MOB and CTree it 
can be deduced that categorizing split variables and binarizing scores has a negative effect 
on the ability to detect the right split variable. To elaborate these effects more in detail the results for all possible combinations of the building blocks are discussed in a full factorial 3-way analysis in the following section.

\textit{Note:} As explained before, the tests are applied as significance tests with 
significance level $\alpha=0.05$. To show that in general the drawn conclusions do not 
change if no significance level is included the same setting is evaluated again without 
any significance level and the results are presented in Appendix~\ref{app:significance}.

\subsection{Full factorial analysis}\label{sec:3way}
To investigate the effects of each of the building blocks separately the most general 
case of the first simulation scenario (stump with one split) where the intercept and 
slope parameter are both varying is considered. In the underlying evaluation all 
possible combinations of the building blocks have been included. The different levels of 
each of the three building blocks are listed in Table~\ref{tab:levels}. A more detailed
overview of all 12 combinations can be found in Appendix~\ref{app:combinations}, 
Table~\ref{tab:combinations_appendix}.

@@

Based on the results displayed in Figure~\ref{fig:3way} it can be stated that binarizing residuals/scores decreases the performance 
as it leads to higher $p$-values corresponding to the true split variable $Z_1$ 
and thus to lower power in the presented settings. 
For a central true split point (left panel) this effect remains at an almost constant level 
regardless of whether possible split variables are categorized or not. However, in case 
of the true split point being closer to the upper margin ($\xi=0.8$, right panel), categorizing
split variables increases the $p$-value of $Z_1$ even more. 
Here, the maximal selection performs best while the linear selection shows its disadvantage in 
detecting abrupt changes. Overall, it depends on the situation whether a linear or a maximal
selection of a split variable is advantageous. But unless one of the breakpoints used for binning possible split variables is close to the true split point (as for $\xi=0$, left panel),
this categorization weakens the performance of a testing strategy.

While the effect of both of these transformations of the data (binarization of 
residuals/scores and categorization of split variables) can be observed
separately in Figure~\ref{fig:3way}, combining them increases the negative impact 
remarkably.
As already shown in Section~\ref{sec:scores} the use of scores vs.\ residuals has a minor 
effect if there is a change in the intercept which is
supported by the small differences between the dashed lines (scores) and the solid
lines (residuals) in Figure~\ref{fig:3way}.

Note that the choice of the two specific settings 
(left panel: true split point $\xi=0$,  effect size $\delta = 0.3$; 
right panel: $\xi=0.8$ requiring a higher effect size $\delta = 1$)
has been made based on the results shown in Figure~\ref{fig:scores_stump} in order to make 
the differences particularly well observable. (Additional evaluations for increasing $\delta$ 
as in Section~\ref{sec:scores} can be found in Appendix~\ref{app:increasingdelta}.)

@@

#par(mar = c(3,5,1,5.7))
#with(d00, interaction.plot(res_scores, interaction(cat, bin), pval_z1, ylim = c(0,1),
#                           ylab = expression("Mean"~p~"-value for"~Z[1])))
#with(d00, interaction.plot(interaction(cat, bin), res_scores, pval_z1))
#with(d00, interaction.plot(cat, interaction(bin, res_scores), pval_z1))
#with(d08, interaction.plot(res_scores, interaction(cat, bin), pval_z1, ylim = c(0,1)))
#with(d08, interaction.plot(interaction(cat, bin), res_scores, pval_z1))
#with(d08, interaction.plot(cat, interaction(bin, res_scores), pval_z1))
@
\caption{\label{fig:3way} Effects of the three building blocks \textit{residuals vs.\ scores},
\textit{binarization}, and \textit{categorization}. All possible combinations have been 
evaluated in two different settings of the first simulation study (stump with one split) 
and their performances are compared based on the mean $p$-values corresponding to the 
true split variable $Z_1$. 
Left panel: true split point $\xi=0$ and effect size $\delta = 0.3$.
Right panel: true split point $\xi=0.8$ and effect size $\delta = 1$.}
\end{figure}

@@

@
\caption{\label{fig:tree_ari}
CTree, MOB, GUIDE, and GUIDE+scores evaluated on the second simulation scenario (tree with two splits), all
applying pre-pruning. Their performances are compared based on the Adjusted Rand Index over
increasing effect size $\delta$ (100 replications of 250 observations per step).
Each panel represents a different value of the true split point $\xi$ ($0$, $0.2$, $0.5$ and 
$0.8$ from left to right).}
\end{figure}

@@

\begin{figure}%[t]
\setkeys{Gin}{width=\linewidth}
<<prune_ari, fig=TRUE, echo=FALSE, height=4, width=8>>=
xyplot(ari ~ delta | xi, groups = ~ test, 
       data = prex, 
       type = "l", lty = c(2,2,2,2,1,1,1,1),
       lwd = 2,
       key =  list(text = list(levels(prex$test)[1:4]),
                   lines = list(lty = c(1,1,1,1), type = "l", pch = 1, cex = 0.8,
                                col = pal[c("CTree", "MOB", "GUIDE+scores", "GUIDE")], lwd = 2)),
       layout= c(4,1),
       scales=list(x=list(at=seq(1,11,2))),
       xlab = expression(delta),
       ylab = "Adjusted Rand Index",
       strip = strip.custom(factor.levels = c(expression(xi==0), 
                                              expression(xi==0.2), 
                                              expression(xi==0.5), 
                                              expression(xi==0.8)),
                                   bg = "gray90"),
       par.settings = par.settings
)
@
\caption{\label{fig:pruning_ari}
CTree, MOB, GUIDE, and GUIDE+scores evaluated on the second simulation scenario (tree), 
once applying pre-pruning (dashed lines) and once post-pruning (solid lines). 
Their performances are compared based on the Adjusted Rand Index over increasing effect 
size $\delta$ (100 replications of 250 observations per step).
Each panel represents a different value of the true split point $\xi$ ($0$, $0.2$, $0.5$ and 
$0.8$ from left to right).}
\end{figure}

@@

For the evaluation of all 12 possible combinations of the presented building blocks
the original testing strategies CTree, MOB, and GUIDE have been applied together
with modified versions of them. The employed versions and the corresponding combination
of building blocks are listed in Table~\ref{tab:combinations_appendix}.
Each suffix indicates a change of the original form of the considered testing strategy.
In particular, it relates to the transformations of the data before applying a statistical
test as represented by the different building blocks. In that way,
``+res'' refers to using only residuals (instead of all scores),
``+scores'' to using all scores (instead of only residuals),
``+cat'' to applying a categorization of split variables,
``+bin'' to applying a binarization of scores/residuals.
